---
title: 'Expression variance: Syngnathus scovelli'
author: "Coley Tosto"
date: "2023-10-10"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r knitsetup, include=FALSE}
knitr::opts_knit$set(root.dir='../',fig_path="./figs/")
```
```{r chunksetup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,out.extra='',fig.pos="H",
                      fig.path = "./figs/",
                      dpi = 300,fig.keep='none',dev='png')
```

```{r loadlibs, message=FALSE}
library(spfTools)
library(viridis)
library(DESeq2)
```

```{r}
#Read in the RDS file that contains expression data
SStxi <- readRDS("data/txi.salmon_SS.RDS")

#The samples file generated for tximport
samples <- read.table("SS_samples.txt", header = TRUE)

#Make sure the conditions are in the samples file as a factor
samples$Sex <- as.factor(samples$Sex)
samples$Organ <- as.factor(samples$Organ)

```


Using counts, calculate mean and SEM for each sex and each tissue type. Maybe try with both abundances and raw counts. 

Perhaps do similar co-expression plots for the four tissue types as well (a 3-D plot, maybe?)

```{r}
#Create an additional column the has the two conditions combined(sex and organ type)
samples$group <- factor(paste0(samples$Sex, samples$Organ))

##Create the DESeq dataset
ddsMF_SS <- DESeqDataSetFromTximport(SStxi,
                                     colData = samples,
                                     design = ~ group)

##Remove the brood pouch samples from the dataset
ddsMF_SS <- ddsMF_SS[, ddsMF_SS$Organ != "BP"]

ddsMF_SS$group <- as.character(ddsMF_SS$group)
ddsMF_SS$group <- as.factor(ddsMF_SS$group)

##Filter the dataset, only keeping rows that have at least 10 reads total
keep <- rowSums(counts(ddsMF_SS)) >= 10 
ddsMF_SS <- ddsMF_SS[keep, ]

```


```{r calc-}

txpt_sum_stats <- data.frame(matrix(data=NA,
                                  nrow=nrow(counts(ddsMF_SS)),
                                  ncol=24))

rownames(txpt_sum_stats) <- rownames(counts(ddsMF_SS))
colnames(txpt_sum_stats)<-c(
  "Bra_Fem_Mean",
  "Bra_Fem_LSEM",
  "Bra_Fem_USEM",
  "Bra_Mal_Mean",
  "Bra_Mal_LSEM",
  "Bra_Mal_USEM",
  "Gon_Fem_Mean",
  "Gon_Fem_LSEM",
  "Gon_Fem_USEM",
  "Gon_Mal_Mean",
  "Gon_Mal_LSEM",
  "Gon_Mal_USEM",
  "Liv_Fem_Mean",
  "Liv_Fem_LSEM",
  "Liv_Fem_USEM",
  "Liv_Mal_Mean",
  "Liv_Mal_LSEM",
  "Liv_Mal_USEM",
  "Ski_Fem_Mean",
  "Ski_Fem_LSEM",
  "Ski_Fem_USEM",
  "Ski_Mal_Mean",
  "Ski_Mal_LSEM",
  "Ski_Mal_USEM"
)

for(tissue in unique(samples$Organ)){
  
  #Pull out the column names that are of interest for this organ
  out_cols <- grep(substring(tissue,1,3),
                   colnames(txpt_sum_stats), 
                   value=TRUE)
  
  #Save the counts only related to this organ into an object
  cnts <- counts(ddsMF_SS)[, colnames(counts(ddsMF_SS)) %in%
                             samples$ID[samples$Organ == tissue]]
  
  for(sex in unique(samples$Sex)){
    
    #Subset the counts even further to only include one sex
    sex_cnts <- cnts[, colnames(cnts) %in%
                        samples$ID[samples$Sex == sex]]
    
    #Calculate the mean for each row (of the log transformed value)
    sex_mean <- rowMeans(log(sex_cnts + 1))
    
    #Calculate the standard error for each row
    sex_sem <- apply(log(sex_cnts + 1), 1, sem)
    
    #Save all of the values for the mean into the corresponding column
    txpt_sum_stats[, grep(paste0(sex,".*_Mean"),
                          out_cols, 
                          value=TRUE)] <- sex_mean
    
    #Calculate the lower end of the CI and save that into the corresponding column
    txpt_sum_stats[, grep(paste0(sex,".*_LSEM"),
                          out_cols, 
                          value=TRUE)] <- txpt_sum_stats[, grep(paste0(sex, ".*_Mean"),
                                                                out_cols, 
                                                                value=TRUE)] - sex_sem
    txpt_sum_stats[, grep(paste0(sex,".*_USEM"), 
                          out_cols,
                          value=TRUE)] <- txpt_sum_stats[, grep(paste0(sex,".*_Mean"),
                                                                out_cols, 
                                                                value=TRUE)] + sex_sem
  }
}
```

```{r readin_DE_results}
#Specify the directory where diff. expression results are located
sbg_path <- "data/"

#Create a list of the files I want
SBG_files <- list.files(sbg_path, pattern = "_res")

#Create an empty list to store my datasets in
SBG_list <- list()

#Create a loop to read in all of the blast results
for(file_name in SBG_files){

  #Read the file and store it in a new object
  file_data <- read.csv(file.path(sbg_path, file_name), 
                        row.names = 1, 
                        header = TRUE)
  
  #Create a new object with a name based on the file name
  sbg_name <- sub("_res.csv$", "", file_name) #Removes the file extension
  SBG_list[[sbg_name]] <- file_data
}

```

```{r varFig, fig.keep='last', fig.height=5, fig.width=7.5}


par(mfrow=c(1, 4))
organs <- c("Brain", "Gonad", "Liver", "Skin")

for(i in 1:4){
  #browser()
  tissue <- organs[i]
  #Pull out the columns corresponding to this tissue specifically
  out_cols <- grep(substring(tissue, 1, 3),
                   colnames(txpt_sum_stats), 
                   value=TRUE)
  
  tmp <- txpt_sum_stats[, out_cols]
  
  
  #Merge the mean and CI data with logFC data
  tmp <- merge(SBG_list[[i]], tmp, by = "row.names", all.x = TRUE)
  
  #Set the intervals used to add color to the plot
  pal <- mako(200,
              alpha=0.5,
              direction = -1)
  intervals <- findInterval(abs(tmp$log2FoldChange),
                            seq(min(abs(tmp$log2FoldChange), na.rm = TRUE),
                                max(abs(tmp$log2FoldChange), na.rm=TRUE),
                                length.out=200),
                            all.inside = T)
  col_test <- pal[intervals]
  
  #Set the max value for the y-axis
  ymax <- max(tmp[, grep("USEM",
                         colnames(tmp))], 
              na.rm = TRUE)

  
  plot(tmp[, grep("Fem_Mean", out_cols, value=TRUE)],
       tmp[, grep("Mal_Mean", out_cols, value=TRUE)],
       xlim = c(0, ymax),
       ylim = c(0, ymax),
       xlab = "log female gene counts",
       ylab = "log male gene counts",
       bty = 'n',
       main = tolower(tissue),
       col = col_test,# "grey", #colors[txpt_sum_stats$SexBiased],
       cex = 0.5,
       pch = 19)
  
  # add bar in female direction
  arrows(x0 = tmp[,grep("Fem_LSEM", out_cols,value=TRUE)],
         y0 = tmp[,grep("Mal_Mean", out_cols,value=TRUE)],
         x1 = tmp[,grep("Fem_USEM", out_cols,value=TRUE)],
         y1 = tmp[,grep("Mal_Mean", out_cols,value=TRUE)],
         length=0,
         col=col_test)
  
  # add bar in male direction
  arrows(y0 = tmp[,grep("Mal_LSEM", out_cols,value=TRUE)],
         x0 = tmp[,grep("Fem_Mean", out_cols,value=TRUE)],
         y1 = tmp[,grep("Mal_USEM", out_cols,value=TRUE)],
         x1 = tmp[,grep("Fem_Mean", out_cols,value=TRUE)],
         length=0,
         col=col_test# colors[txpt_sum_stats$SexBiased])
  )
}

par(xpd=TRUE)

legend_image <- as.raster(matrix(mako(200, alpha=0.5), ncol=1))

rasterImage(legend_image, 14.5, 10, 15.5, 15)                       ## the gradient

text(round(min(abs(res$log2FoldChange), na.rm = TRUE),1),
     x=16, y=10, cex=0.6)
text(round(max(abs(res$log2FoldChange), na.rm=TRUE),1),
     x=16, y=15, cex=0.6)
text('log2FC', x=15, y=15.5, cex=.6, font=2)          ## legend title



```

It's interesting to see all four of the organs, however, I only have enough data surrounding Tajima's D for the skin, gonads, and liver. Therefore, I just want to plot those three. 

I may also play around with denoting sex-bias in the same color scheme that we have before to keep some cohesion across the different figures

```{r varFig, fig.keep='last', fig.height=5, fig.width=7.5}
pdf("docs/figs/Fig_var_expression.pdf", width = 10, height=4)

par(mfrow=c(1,3), 
    oma=c(4,4,2,8), 
    mar=c(1,1,1,0))

for(i in 2:4){
  
  tissue <- organs[i]
  #Pull out the columns corresponding to this tissue specifically
  out_cols <- grep(substring(tissue, 1, 3),
                   colnames(txpt_sum_stats), 
                   value=TRUE)
  
  tmp <- txpt_sum_stats[, out_cols]
  
  
  #Merge the mean and CI data with logFC data
  tmp <- merge(SBG_list[[i]], tmp, by = "row.names", all.x = TRUE)

  
  plot(tmp[, grep("Fem_Mean", out_cols, value=TRUE)],
       tmp[, grep("Mal_Mean", out_cols, value=TRUE)],
       xlim = c(0, 15),
       ylim = c(0, 15),
       axes = FALSE,
       bty = 'n',
       cex.main = 2,
       col = ifelse(tmp$log2FoldChange > 2 & tmp$padj < 0.05,
                    "#beaed475",
                    ifelse(tmp$log2FoldChange < -2 & tmp$padj < 0.05,
                           "#7fc97f75",
                           "#A9A9A975")),#colors[txpt_sum_stats$SexBiased],
       cex = 0.5,
       pch = 19)
  
  axis(1,
       lwd = 2,cex = 2, cex.axis=1.5, las=1)
  axis(2,
       lwd=2,cex=2, cex.axis=1.5, las=1)
  
  mtext(tissue, cex=1.5,outer=FALSE, line=-1)
  
  # add bar in female direction
  arrows(x0 = tmp[,grep("Fem_LSEM", out_cols,value=TRUE)],
         y0 = tmp[,grep("Mal_Mean", out_cols,value=TRUE)],
         x1 = tmp[,grep("Fem_USEM", out_cols,value=TRUE)],
         y1 = tmp[,grep("Mal_Mean", out_cols,value=TRUE)],
         length=0,
         col=ifelse(tmp$log2FoldChange > 2 & tmp$padj < 0.05,
                    "#beaed450",
                    ifelse(tmp$log2FoldChange < -2 & tmp$padj < 0.05,
                           "#7fc97f50",
                           "#A9A9A950")))
  
  # add bar in male direction
  arrows(y0 = tmp[,grep("Mal_LSEM", out_cols,value=TRUE)],
         x0 = tmp[,grep("Fem_Mean", out_cols,value=TRUE)],
         y1 = tmp[,grep("Mal_USEM", out_cols,value=TRUE)],
         x1 = tmp[,grep("Fem_Mean", out_cols,value=TRUE)],
         length=0,
         col=ifelse(tmp$log2FoldChange > 2 & tmp$padj < 0.05,
                    "#beaed450",
                    ifelse(tmp$log2FoldChange < -2 & tmp$padj < 0.05,
                           "#7fc97f50",
                           "#A9A9A950")))
  
}


# add x axis label
mtext("log(Female gene counts)",side = 1,cex=1.5, outer=TRUE, line=2)

# add y axis label
mtext("log(Male gene counts)",side = 2,cex=1.5, outer=TRUE, line=2.25)

# add legend
spfTools::outer_legend("right",
                       legend=c("Female\nbiased",
                                "Male\nbiased",
                                "Unbiased"),
                       pt.bg=c("#7fc97f", "#beaed4", "#A9A9A9"),
                       pch=22,
                       bty='n',
                       ncol=1,
                       cex=2,
                       y.intersp = 1.5,
                       pt.cex=2)


dev.off()

```

```{r varFig, fig.keep='last', fig.height=5, fig.width=7.5}
pdf("docs/figs/Fig_var_expression_all.pdf", width = 10, height=4)

par(mfrow=c(1,4), 
    oma=c(4,4,2,8), 
    mar=c(1,1,1,0))

for(i in 1:4){
  
  tissue <- organs[i]
  #Pull out the columns corresponding to this tissue specifically
  out_cols <- grep(substring(tissue, 1, 3),
                   colnames(txpt_sum_stats), 
                   value=TRUE)
  
  tmp <- txpt_sum_stats[, out_cols]
  
  
  #Merge the mean and CI data with logFC data
  tmp <- merge(SBG_list[[i]], tmp, by = "row.names", all.x = TRUE)

  
  plot(tmp[, grep("Fem_Mean", out_cols, value=TRUE)],
       tmp[, grep("Mal_Mean", out_cols, value=TRUE)],
       xlim = c(0, 15),
       ylim = c(0, 15),
       axes = FALSE,
       bty = 'n',
       cex.main = 2,
       col = ifelse(tmp$log2FoldChange > 2 & tmp$padj < 0.05,
                    "#beaed475",
                    ifelse(tmp$log2FoldChange < -2 & tmp$padj < 0.05,
                           "#7fc97f75",
                           "#A9A9A975")),
       cex = 0.5,
       pch = 19)
  
  axis(1,
       lwd = 2,cex = 2, cex.axis=1.5, las=1)
  axis(2,
       lwd=2,cex=2, cex.axis=1.5, las=1)
  
  mtext(tissue, cex=1.5,outer=FALSE, line=-1)
  
  # add bar in female direction
  arrows(x0 = tmp[,grep("Fem_LSEM", out_cols,value=TRUE)],
         y0 = tmp[,grep("Mal_Mean", out_cols,value=TRUE)],
         x1 = tmp[,grep("Fem_USEM", out_cols,value=TRUE)],
         y1 = tmp[,grep("Mal_Mean", out_cols,value=TRUE)],
         length=0,
         col=ifelse(tmp$log2FoldChange > 2 & tmp$padj < 0.05,
                    "#beaed450",
                    ifelse(tmp$log2FoldChange < -2 & tmp$padj < 0.05,
                           "#7fc97f50",
                           "#A9A9A950")))
  
  # add bar in male direction
  arrows(y0 = tmp[,grep("Mal_LSEM", out_cols,value=TRUE)],
         x0 = tmp[,grep("Fem_Mean", out_cols,value=TRUE)],
         y1 = tmp[,grep("Mal_USEM", out_cols,value=TRUE)],
         x1 = tmp[,grep("Fem_Mean", out_cols,value=TRUE)],
         length=0,
         col=ifelse(tmp$log2FoldChange > 2 & tmp$padj < 0.05,
                    "#beaed450",
                    ifelse(tmp$log2FoldChange < -2 & tmp$padj < 0.05,
                           "#7fc97f50",
                           "#A9A9A950")))
  
}


# add x axis label
mtext("log(Female gene counts)",side = 1,cex=1.5, outer=TRUE, line=2)

# add y axis label
mtext("log(Male gene counts)",side = 2,cex=1.5, outer=TRUE, line=2.25)

# add legend
spfTools::outer_legend("right",
                       legend=c("Female\nbiased",
                                "Male\nbiased",
                                "Unbiased"),
                       pt.bg=c("#7fc97f", "#beaed4", "#A9A9A9"),
                       pch=22,
                       bty='n',
                       ncol=1,
                       cex=2,
                       y.intersp = 1.5,
                       pt.cex=2)


dev.off()

```